A Practical Optional Type System for Clojure

### 2012-10-21 dissertation/intro.tex
2nd page of revisions from rowan

Clojure has attracted wide-spread users in part by concentrating on pragmatism.
Performance is a key feature, for example the JVM implementation of Clojure
offers ways to access Java-like speed for certain operations.
Also, Clojure's extensive host interoperability offers Clojure programmers
access to existing libraries for their platform, such as the extensive Java library ecosystem.
By coupling pragmatic necessities with elegant features like Lisp-style macros, functional programming,
and immutability by default, Clojure is a compelling general purpose programming language.

@@

He highlighted a strong desire for a type system for several reasons.
Firstly, to verify the correctness of the library.
Without a static type system, it is a significant task
to verify such a implementation correct due to heavy use
of higher-order functions.
Secondly, to aid him while writing the library.

@@

There are many concepts associated with \emph{types} and \emph{type systems} in both the
literature and informal discourse.
A programmer who uses dynamically-typed languages may have a drastically different notion
of what a type is than, say, a programmer preferring languages with advanced static type systems.
There is some debate as to whether optional static type systems like Typed Clojure
can even be called a type system. We choose to
follow the terminology of Pfenning~\cite{Pfe08} and Reynolds~\cite{Rey01},
where such optional type systems are \emph{extrinsic} type systems, and more
traditional type systems are \emph{intrinsic} type systems.
This distinction has a long history, originating in work in the

An ordinary static type system is used to check whether programs are basically
meaningful. Pfenning and Reynolds call these type systems \emph{intrinsic}. A language with an intrinsic
static type system has run-time semantics that depends on the types of associated
variables and expressions during type checking.
For example, C, Java, and ML have intrinsic types.
This means programs written in these languages must pass the type checker before being run.

A static type system is \emph{extrinsic} when runtime semantics
does not depend on the type system. In other words, passing a static type checker
is not essential to running programs. A dynamically typed language can be viewed
as having a trivial static type system that supports exact one type,
a view strongly advocated by Harper~\cite{Har12}
and common in the literature on static types (for example, Pierce~\cite{Pie02}).

@@

For example, the Java expressions \emph{1 + 2 / 3} is written in Lisp pseudocode \lstinline|(/ (+ 1 2) 3)|
and \emph{numberCrunch(1, 2)} written \lstinline|(numberCrunch 1 2)|.

@@

\begin{itemize}
  \item Prefixing \lstinline|:| to a symbol defines a \emph{keyword}, often used for map keys. eg. \lstinline|:my-keyword|.
  \item Square brackets delimit vector literals. eg. \lstinline|[1 2]| is a 2 place vector.
  \item Curly brackets define map literals. eg. \lstinline|{:a 1 :b 2}| is a map from 
        \lstinline|:a| to \lstinline|1| and \lstinline|:b| to \lstinline|2|.
  \item Commas are always optional, and treated as whitespace.
\end{itemize}

@@

At this point, it is worth understanding Clojure's namespacing feature.
Clojure code is always executed in a \emph{namespace}, and each file of Clojure code should 
have a \lstinline|ns| declaration with the namespace name and its dependencies,
which switches the current namespace and executes the given dependency commands.
There is one special namespace, \lstinline|clojure.core| which is
loaded with every namespace, implicitly ``referring'' all its vars in the namespace.
For example, \lstinline|ns| refers to the var \lstinline|clojure.core/ns|,
similarly \lstinline|println| refers to \lstinline|clojure.core/println|.

The example in listing \ref{lst:helloworld} declares a dependency to 
\lstinline|typed.core|, Typed Clojure's main namespace. It also refers the var \lstinline|typed.core/check-ns|
into scope\footnote{\lstinline|check-ns| is the top level function for type checking a namespace}.
Other than this dependency, this is identical to the untyped \emph{Hello world}.

More complex code require extra annotations to type check:

\begin{lstlisting}[caption=Annotating vars in Typed Clojure]
(ns typed.test.collatz
  (:require [typed.core :refer [check-ns ann]]))

(ann collatz [Number -> Number])
(defn collatz [n]
  (cond
    (= 1 n) 
      1
    (and (integer? n) 
         (even? n)) 
      (collatz (/ n 2))
    :else 
      (collatz (inc (* 3 n)))))
\end{lstlisting}
\footnote{Example adapted from Tobin-Hochstadt~\cite{Tob10}}

In this example, we define a new var \lstinline|typed.test.collatz/collatz|. Typed Clojure requires all 
used vars to be annotated. Here \lstinline|typed.core/ann| annotates \lstinline|typed.test.collatz/collatz|
to be a function from \lstinline{java.lang.Number} to 
\lstinline{java.lang.Number}\footnote{All Classes in the \lstinline|java.lang| package
are automatically imported in every Clojure namespace (the equivalent of Java's \lstinline|import java.lang.*;|).}.

@@

We can annotate datatype and protocol definitions similarly.

@@

\lstinline|clojure.core/defprotocol| defines a new Clojure protocol\footnote{See http://clojure.org/protocols for a full description of protocols}
with a set of methods. \lstinline|typed.core/ann-protocol| annotates a protocol with the types of its methods.
In this example, we define a protocol \lstinline|typed.test.person/Age| with an \lstinline|age| method.
The call to \lstinline|clojure.core/defprotocol| also defines a new var \lstinline|typed.test.person/age|, a first-class function
wrapping the \lstinline|age| method, but taking the target Object as the first parameter. The
type signature provided with \lstinline|typed.core/ann-protocol|, here \lstinline|[typed.test.person/Age -> typed.core/AnyInteger]|, 
is for this function.

Invocations of \lstinline|clojure.core/defprotocol| are currently not able to be type checked
and are ignored by Typed Clojure by passing them to \lstinline|typed.core/tc-ignore|.

\lstinline|clojure.core/deftype|
defines a new Clojure datatype\footnote{See http://clojure.org/datatypes for a full description of datatypes}
in the current namespace with a number of fields and methods. 
\lstinline|typed.core/ann-datatype| annotates a datatype with its field types.
In this example, we create a datatype \lstinline|typed.test.person.Person|
with fields \lstinline|name| and \lstinline|age| and implement the \lstinline|age|
method from protocol \lstinline|typed.test.person/Age|.

Java constructors are invoked in Clojure by suffixing the Class we want to instantiate with a dot.
Datatypes are implemented as Java Classes with immutable fields (by default) and a single constructor, taking as arguments its fields 
in the order they are passed to \lstinline|deftype|
\footnote{When unambiguous, I omit the qualifying namespace/package for the remainder of the chapter.}.
\lstinline|(Person. "Lucy" 34)| constructs a new \lstinline|Person|
instance, setting the fields to their corresponding positional arguments.
Typed Clojure checks the datatype constructor to be the equivalent of 
\lstinline|[String AnyInteger -> Person]|.

Finally, Typed Clojure checks invocations of Protocol methods. It infers \lstinline|Person|
is an instance of \lstinline|Age| from the datatype definition, therefore \lstinline|(age (Person. "Lucy" 34))| is type-safe.

@@

Typed Clojure supports F-bounded polymorphism, first introduced by Canning, Cook, Hill and Olthoff~\cite{CCHOM89}. 
F-bounded polymorphism is an extension of bounded polymorphism, where polymorphic type variables
can be restricted by \emph{bounds}.
In particular, F-bounded polymorphism allows type variable bounds to recursively refer to the
variable being bounded. Typed Clojure supports upper and lower type variable bounds.

Typed Clojure parameterises some of Clojure's data structures. For example,
the interface behind Clojure's \emph{seq} abstraction \lstinline|clojure.lang.Seqable| has one 
covariant parameter\footnote{\lstinline|(Seqable Integer)| being a subtype of \lstinline|(Seqable Number)|
because Integer is a subtype of Number.}.

@@

\lstinline|(U nil (Seqable x))| is a common type in Typed Clojure, read as the union
of the type \lstinline|nil| and the type \lstinline|(Seqable x)|.
The vast majority of types for collection processing functions in the Clojure core library feature
it as an input type, where passing \lstinline|nil| either has some special behaviour 
or is synonymous with passing an empty \lstinline|Seqable|.

@@

A novel feature of Clojure compared to other dialects of Lisp (including Racket)
is an emphasis on key-value paired, unordered \emph{maps}.
Where particularly object-oriented languages reach for objects, Clojure
utilises maps. Clojure provides a map literal using curly braces. For example,
\lstinline|{:a 1, :b 2}| is a map value with two key-value entries: from keyword key \lstinline|:a|
to value \lstinline|1|, and keyword key \lstinline|:b| to value \lstinline|2|. Note that commas are always
whitespace in Clojure and are included occasionally for readability.

Typed Clojure provides a heterogeneous map type which captures this common
``maps as objects'' pattern. A heterogeneous map type has only \emph{positive}
information on the types of key-value entries. In other words, it conveys
whether a particular key is present, but not whether it is absent.
The implications of this is discussed in section \ref{ref:designhmap}.
Heterogeneous maps only support keys that are singleton Keyword types. This restriction is reflected
in the syntax for defining heterogeneous map types.

@@

Heterogeneous vector and seq types are also provided.

@@

Functions in Clojure are multi-arity, which means a function
can be defined with several function bodies (or \emph{arities})
and which arity is executed depends on the number of arguments passed
to the function. A function can have any number of arities with fixed parameters, and at most one arity with
variable-parameters. Each arity must have a unique number of
parameters and have a lower number of parameters than the ``variable arity'', if present.

>>>>

Clojure has attracted wide-spread users in part by concentrating on pragmatism.
Performance is a key feature, for example the JVM implementation of Clojure
offers ways to access Java-like speed for certain operations.
Also, Clojure's extensive host interoperability offers Clojure programmers
access to existing libraries for their platform, such as the vast Java library ecosystem.
By coupling pragmatic necessities with elegant features like Lisp-style macros, functional programming,
and immutability by default, Clojure is a compelling general purpose programming language.

@@

He highlighted a strong desire for a type system for several reasons.
Firstly, to verify the correctness of the library.
Without a static type system, it is a significant task
to verify such an implementation as correct due to heavy use
of higher-order functions.
Secondly, to aid him while writing the library.

@@

There are many concepts associated with \emph{types} and \emph{type systems} in both the
literature and informal discourse.
A programmer who uses dynamically-typed languages may have a drastically different notion
of what a type is than, say, a programmer preferring languages with advanced static type systems.
There is some debate as to whether optional static type systems like Typed Clojure
can even be called a type system. We choose to
follow the terminology of Pfenning~\cite{Pfe08} and Reynolds~\cite{Rey01},
where such optional type systems are \emph{extrinsic} type systems, and more
traditional type systems are \emph{intrinsic} type systems.
This distinction has a long history, originating in work in the $\lambda$-calculus 
by Church \footnote{CHURCH FROM ROWANS DISSERTATION} and Curry \footnote{CURRY FROM ROWANS DISSERTATION},
leading to intrinsic types and extrinsic types also being called Church types and
Curry types often in the literature.

An ordinary static type system is used to check whether programs are basically
meaningful. Pfenning and Reynolds call these type systems \emph{intrinsic}. A language with an intrinsic
static type system has a run-time semantics that depends on the types associated
with variables and expressions during type checking.
For example, C, Java, ML, and Haskell have intrinsic types.
This means programs written in these languages must pass the type checker before being run.

A static type system is \emph{extrinsic} when runtime semantics
does not depend on a static type system. In other words, passing a static type checker
is not essential to running programs. A dynamically typed language can be viewed
as having a trivial static type system that supports exactly one type,
a view advocated by Harper~\cite{Har12}
and common in the literature on static types (for example, Pierce~\cite{Pie02}).

@@

For example, the Java expressions \emph{(1 + 2) / 3} is written in Lisp pseudocode \lstinline|(/ (+ 1 2) 3)|
and \emph{numberCrunch(1, 2)} written \lstinline|(numberCrunch 1 2)|.

@@

\begin{itemize}
  \item Prefixing \lstinline|:| to a symbol defines a \emph{keyword}, often used for map keys. eg. \lstinline|:my-keyword|.
  \item Square brackets delimit vector literals. eg. \lstinline|[1 2]| is a 2 place vector.
  \item Curly brackets define map literals. eg. \lstinline|{:a 1 :b 2}| is a map from 
        \lstinline|:a| to \lstinline|1| and \lstinline|:b| to \lstinline|2|.
  \item Commas are always optional and treated as whitespace, but often used to show the intended structure.
\end{itemize}

@@

At this point, it is worth understanding Clojure's namespacing feature.
Clojure code is always executed in a \emph{namespace}, and each file of Clojure code should 
have a \lstinline|ns| declaration with the namespace name and its dependencies,
which switches the current namespace and executes the given dependency commands.
There is one special namespace, \lstinline|clojure.core| which is
loaded with every namespace, implicitly ``referring'' all its vars in the namespace.
For example, \lstinline|ns| refers to the var \lstinline|clojure.core/ns|,
similarly \lstinline|println| refers to \lstinline|clojure.core/println|
(vars are global bindings in Clojure).

The example in listing \ref{lst:helloworld} declares a dependency to 
\lstinline|typed.core|, Typed Clojure's main namespace. It also refers the var \lstinline|typed.core/check-ns|
into scope (\lstinline|check-ns| is the top level function for type checking a namespace).
Other than this dependency, this is identical to the untyped \emph{Hello world}.

More complex code may require extra annotations to type check:

\begin{lstlisting}[caption=Annotating vars in Typed Clojure (adapted from a Typed Scheme/Racket example by Tobin-Hochstadt~\cite{Tob10})]
(ns typed.test.collatz
  (:require [typed.core :refer [check-ns ann]]))

(ann collatz [Number -> Number])
(defn collatz [n]
  (cond
    (= 1 n) 
      1
    (and (integer? n) 
         (even? n)) 
      (collatz (/ n 2))
    :else 
      (collatz (inc (* 3 n)))))
\end{lstlisting}

(\lstinline|(defn [...] body)| expands to \lstinline|(def (fn [...] body))|
where \lstinline|def| defines a new var and \lstinline|fn| creates a function value).

In this example, we define a new var \lstinline|collatz|
(when unambiguous, I omit the qualifying namespace/package for the remainder of the chapter).
Typed Clojure requires all 
used vars to be annotated. Here \lstinline|typed.core/ann|, a Typed Clojure macro for annotating vars, 
annotates \lstinline|collatz| to be a function from \lstinline|Number| to 
\lstinline|Number| (\lstinline|Number| refers to \lstinline|java.lang.Number|,
due to every Clojure namespace implicitly importing all Classes in the \lstinline|java.lang| package,
the equivalent of Java's \lstinline|import java.lang.*;|).

@@

As well as \lstinline|def| and \lstinline|defn| definitions,
Clojure programmers typically include datatype and protocol
definitions. 
Protocols are similar to Java interfaces\footnote{JAVA INTERFACE REFERENCE} and datatypes to classes implementing interfaces.
We can annotate datatype and protocol definitions similarly.

@@

\lstinline|defprotocol| defines a new Clojure protocol\footnote{See http://clojure.org/protocols for a full description of protocols.}
with a set of methods. \lstinline|ann-protocol| is a Typed Clojure macro for annotating 
a protocol with the types of its methods.
In this example, we define a protocol \lstinline|Age| with an \lstinline|age| method,
which is really a first-class function taking the target object as the first parameter. The
type signature provided with \lstinline|ann-protocol|, here \lstinline|[Age -> AnyInteger]|
which is the type of the function taking an \lstinline|Age| and returning an \lstinline|AnyInteger|,
is for this function.

Invocations of \lstinline|defprotocol| are currently not able to be type checked
and are ignored by Typed Clojure by passing them to \lstinline|typed.core/tc-ignore|.

\lstinline|deftype|
defines a new Clojure datatype\footnote{See http://clojure.org/datatypes for a full description of datatypes.}
in the current namespace with a number of fields and methods. 
\lstinline|typed.core/ann-datatype| is a Typed Clojure form for annotating datatypes, including its field types.
In this example, we create a datatype \lstinline|typed.test.person.Person|
(datatype defintions generate a Java Class, where the current namespace is used as a starting point for its
qualifying package)
with fields \lstinline|name| and \lstinline|age| and 
extend the \lstinline|Age| protocol by implementing the \lstinline|age| method.

Java constructors are invoked in Clojure by suffixing the Class we want to instantiate with a dot.
Datatypes are implemented as Java Classes with immutable fields (by default) and a single constructor, taking as arguments its fields 
in the order they are passed to \lstinline|deftype|
\lstinline|(Person. "Lucy" 34)| constructs a new \lstinline|Person|
instance, setting the fields to their corresponding positional arguments.
Typed Clojure checks the datatype constructor to be the equivalent of 
\lstinline|[String AnyInteger -> Person]|.

Finally, Typed Clojure checks invocations of Protocol methods. It infers \lstinline|Person|
is an instance of \lstinline|Age| from the datatype definition, therefore \lstinline|(age (Person. "Lucy" 34))| is type-safe.
Since generally we only need to annotate definitions and not uses, the number of annotations is relatively small.

@@

Typed Clojure supports F-bounded polymorphism, first introduced by Canning, Cook, Hill and Olthoff~\cite{CCHOM89}. 
F-bounded polymorphism is an extension of bounded polymorphism, where polymorphic type variables
can be restricted by \emph{bounds}.
In particular, F-bounded polymorphism allows type variable bounds to recursively refer to the
variable being bounded. Typed Clojure supports upper and lower type variable bounds.
\footnote{F-BOUND EXAMPLE NEEDED}

For accurate type checking, Typed Clojure parameterises some of Clojure's data structures. For example,
the interface behind Clojure's \emph{seq} abstraction for sequences \lstinline|clojure.lang.Seqable| has one 
covariant parameter\footnote{Covariant means that \lstinline|(Seqable Integer)| is a subtype of \lstinline|(Seqable Number)|
because Integer is a subtype of Number.}.

@@

\lstinline|(U nil (Seqable x))| is a common type in Typed Clojure, read as the union
of the type \lstinline|nil| and the type \lstinline|(Seqable x)|.
The vast majority of types for collection processing functions in the Clojure core library feature
it as an input type, where passing \lstinline|nil| either has some special behaviour 
or is synonymous with passing an empty \lstinline|Seqable|.
The ability to specify types that explicitly include or exclude \lstinline|nil| is one of the strengths
of Typed Clojure, and an aspect where it is more expressive than standard type systems like that for Java.

@@

Clojure emphasises key-value paired, unordered \emph{maps} for many uses where other languages might use objects
Clojure provides a map literal using curly braces. For example,
\lstinline|{:a 1, :b 2}| is a map value with two key-value entries: from keyword key \lstinline|:a|
to value \lstinline|1|, and keyword key \lstinline|:b| to value \lstinline|2|. Note that commas are always
whitespace in Clojure and are included occasionally for readability.

Typed Clojure provides heterogeneous map types which captures this common
``maps as objects'' pattern. A heterogeneous map type has only \emph{positive}
information on the types of key-value entries. In other words, it conveys
whether a particular key is present, but not whether it is absent.
The implications of this is discussed in section \ref{ref:designhmap}.
Heterogeneous maps only support keys that are singleton Keyword types. This restriction is reflected
in the syntax for defining heterogeneous map types.

@@

Heterogeneous vector and seq types are also provided and work similarly,
except for their lack of keys like maps, hence are just a sequence of types.

@@

Functions in Clojure are multi-arity (a very commonly used feature) which means a function
can be defined with several sets of function arguments and bodies (or \emph{arities})
and which arity is executed depends on the number of arguments passed
to the function. A function can have any number of arities with fixed parameters, and at most one arity with
variable-parameters. Each arity must have a unique number of
parameters and have a lower number of parameters than the ``variable arity'', if present.
